Type Refinements for the Working Class

### 2015-11-12 paper.tex
Fix all chktex warnings

\newif\ifcolored
\coloredtrue

@@

\ifcolored
  \def\IModeColorName{MidnightBlue}
  \def\OModeColorName{Maroon}
\else
  \def\IModeColorName{Black}
  \def\OModeColorName{Black}
\fi

@@

\newcommand\Refines[3]{\IMode{#1}\parallel \IMode{#2} \sqsubset \OMode{#3}}
\newcommand\GRefines[2]{\IMode{#1}\sqsubset\OMode{#2}}
\newcommand\RefinesCtx[2]{\IMode{#1}\sqsubset^\star\OMode{#2}}
\newcommand\EqRefines[4]{\IMode{#1}\parallel\IMode{#2}=\IMode{#3}\sqsubset\OMode{#4}}

@@

\newcommand\RMemEq[4]{\IMode{#1}\parallel \IMode{#2}=\IMode{#3}\in \IMode{#4}}
\newcommand\HRMemEq[5]{\IMode{#1}\parallel \IMode{#2} \vDash \IMode{#3}=\IMode{#4}\in \IMode{#5}}

@@

There are two conflicting views which bedevil any discussion of the nature of
type theory. First, there is the notion of type theory as an extension or
generalization of universal algebra to support interdependency of sorts and
operations, possibly subject to an arbitrary equational theory
\cite{cartmell:1986, dybjer:1996}; we will call this \emph{formal type theory}.
Typing, in such a setting, is a mere matter of grammar and is nearly always
decidable. In hindsight, we may observe that this is the sort of type theory
which Martin-L\"of first proposed in 1972 \cite{martin-lof:1972}, even if we
will admit that this was not the intention at the time. A model for such a type
theory is usually given by interpreting the types or sorts as presheaves or
sheaves over contexts of hypotheses, and as such, a proof theoretic
interpretation of the hypothetical judgment is inevitable.

Secondly, there is the view of type theory as semi-formal theory of
constructions for the Brouwer-Heyting-Kolmogorov interpretation of
intuitionistic mathematical language, which we will call \emph{behavioral} or
\emph{semantic type theory}. The most widely known development of this program
is Martin-L\"of's 1979 ``extensional'' type theory \cite{martin-lof:1979,
martin-lof:1984}, but we must give priority to Dana Scott for inventing this
line of research in 1970 with his prophetic report, \emph{Constructive
Validity} \cite{scott:1970}. More recently, behavioral type theory has been
developed much further in the Nuprl family \cite{constable:1986} of proof
assistants, including MetaPRL \cite{hickey:2003} and JonPRL \cite{jonprl:2015}.

@@

The types of the semantic theory can then be said to \emph{refine} the types of
the syntactic theory \cite{harper-davies:2014, harper-duff:2015, harper:2016},
both by placing restrictions on membership and by coarsening equivalence. Thus
far, all developments of behavioral type theory have been built on a
\emph{unityped} syntactic framework, and so the relation to type refinements
has been difficult to see. In this paper, we contribute a full theory of
behavioral refinements over multi-sorted abstract binding trees, a simple
formal type theory \cite{harper:2016, sterling-morrison:2015}; this hybrid
system allows the deployment of a Nuprl-style type theory over any signature of
sorts and operators.

@@

See \cite{sterling-morrison:2015} for the development of abstract binding trees
with symbols. \todo[inline]{give a brief description of the framework, and present its
rules.}

@@

\ifcolored
  In this paper, we hint at the \emph{modes} of judgments and assertions
  \cite{harper:2016} using colors, marking inputs with
  $\IMode{\texttt{blue}}$ and outputs with $\OMode{\texttt{red}}$. As a rule
  of thumb, inputs are things which are supplied when checking the
  correctness of a judgment, and outputs are things which are synthesized in
  the process.
\fi

@@

  Intuitively, the general judgment $\GenJ{\Omega}{\mathcal{J}}$ is evident
  when the family of judgments $\mathcal{J}$ shall forevermore have a section
  (i.e. for any substitution $\Member{\omega}{\square\Omega}$,
  $\mathcal{J}(\omega)$ shall be evident). Recall that metavariables have
  \emph{valences} rather than sorts; as such, the general judgment allows us to
  quantify over terms of higher type.  However, for clarity we will often write
  $\GenJ{\MV{m}:\sigma}{\mathcal{J}}$ rather than
  $\GenJ{\MV{m}:\MkValence{\cdot}{\cdot}{\sigma}}{\mathcal{J}}$; likewise, in
  the body of $\mathcal{J}$ we will simply write $\MV{m}$ rather than the more
  proper $\MApp{\MV{m}}{\cdot}{\cdot}$.

@@

  In other words, the hypothetical judgment
  $\HypJ{\mathcal{J}_2}{\mathcal{J}_1}$ is evident when there shall forevermore
  be an effective transformation of evidence for the antecedent into evidence
  for the consequent. It is important to keep in mind that this is a
  \emph{semantic} consequence and therefore expresses admissibilities, in
  contrast to the \emph{logical} consequence which expresses derivabilities
  \cite{harper:2016}.  We will not have any need for the latter.
\end{definition}

@@

We will require two judgments, defined mutually: $\RefinesCtx{\Phi}{\Upsilon}$,
which means that $\Phi$ is a symbol context which refines the symbol context
$\Upsilon$ (presupposing $\IsSymCtx{\Upsilon}$), and parametric refinement
$\Refines{\Phi}{\phi}{\tau}$, which shall presuppose
$\RefinesCtx{\Phi}{\Upsilon}$ and $\IsSort{\tau}$).

@@

Two refinements are equal when they denote the same PER. That is, will define
the refinement equality judgment $\EqRefines{\Phi}{\phi}{\psi}{\tau}$ as
follows, presupposing $\RefinesCtx{\Phi}{\Upsilon}$,
$\Refines{\Phi}{\phi}{\tau}$ and $\Refines{\Phi}{\psi}{\tau}$:
\begin{definition}
  To know $\EqRefines{\Phi}{\phi}{\psi}{\tau}$ is to know, for any substitution
  $\Of{\rho}{\Upsilon\hookrightarrow\OMode{\Upsilon'}}$, and for any $M, N$
  such that $\IsAbt{\cdot}{\Upsilon'}{\cdot}{M}{\tau}$ and
  $\IsAbt{\cdot}{\Upsilon'}{\cdot}{N}{\tau}$,
  $\IMode{M}\sim^\rho_\phi\IMode{N}$ if and only if
  $\IMode{M}\sim^\rho_\psi\IMode{N}$.
\end{definition}

>>>>

\newif\ifcolored%
\coloredtrue%

@@

\ifcolored%
  \def\IModeColorName{MidnightBlue}
  \def\OModeColorName{Maroon}
\else
  \def\IModeColorName{Black}
  \def\OModeColorName{Black}
\fi

@@

\newcommand\Refines[3]{\IMode{#1}\parallel\IMode{#2}\sqsubset\OMode{#3}}
\newcommand\GRefines[2]{\IMode{#1}\sqsubset\OMode{#2}}
\newcommand\RefinesCtx[2]{\IMode{#1}\sqsubset^\star\OMode{#2}}
\newcommand\EqRefines[4]{\IMode{#1}\parallel\IMode{#2}=\IMode{#3}\sqsubset\OMode{#4}}

@@

\newcommand\RMemEq[4]{\IMode{#1}\parallel\IMode{#2}=\IMode{#3}\in\IMode{#4}}
\newcommand\HRMemEq[5]{\IMode{#1}\parallel\IMode{#2}\vDash\IMode{#3}=\IMode{#4}\in\IMode{#5}}

@@

There are two conflicting views which bedevil any discussion of the nature of
type theory. First, there is the notion of type theory as an extension or
generalization of universal algebra to support interdependency of sorts and
operations, possibly subject to an arbitrary equational theory
\cite{cartmell:1986, dybjer:1996}; we will call this \emph{formal type theory}.
Typing, in such a setting, is a mere matter of grammar and is nearly always
decidable. In hindsight, we may observe that this is the sort of type theory
which Martin-L\"of first proposed in 1972~\cite{martin-lof:1972}, even if we
will admit that this was not the intention at the time. A model for such a type
theory is usually given by interpreting the types or sorts as presheaves or
sheaves over contexts of hypotheses, and as such, a proof theoretic
interpretation of the hypothetical judgment is inevitable.

Secondly, there is the view of type theory as semi-formal theory of
constructions for the Brouwer-Heyting-Kolmogorov interpretation of
intuitionistic mathematical language, which we will call \emph{behavioral} or
\emph{semantic type theory}. The most widely known development of this program
is Martin-L\"of's 1979 ``extensional'' type theory~\cite{martin-lof:1979,
martin-lof:1984}, but we must give priority to Dana Scott for inventing this
line of research in 1970 with his prophetic report, \emph{Constructive
Validity}~\cite{scott:1970}. More recently, behavioral type theory has been
developed much further in the Nuprl family~\cite{constable:1986} of proof
assistants, including MetaPRL~\cite{hickey:2003} and JonPRL~\cite{jonprl:2015}.

@@

The types of the semantic theory can then be said to \emph{refine} the types of
the syntactic theory~\cite{harper-davies:2014, harper-duff:2015, harper:2016},
both by placing restrictions on membership and by coarsening equivalence. Thus
far, all developments of behavioral type theory have been built on a
\emph{unityped} syntactic framework, and so the relation to type refinements
has been difficult to see. In this paper, we contribute a full theory of
behavioral refinements over multi-sorted abstract binding trees, a simple
formal type theory~\cite{harper:2016, sterling-morrison:2015}; this hybrid
system allows the deployment of a Nuprl-style type theory over any signature of
sorts and operators.

@@

See~\cite{sterling-morrison:2015} for the development of abstract binding trees
with symbols. \todo[inline]{give a brief description of the framework, and present its
rules.}

@@

\ifcolored%
In this paper, we hint at the \emph{modes} of judgments and
assertions~\cite{harper:2016} using colors, marking inputs with
$\IMode{\texttt{blue}}$ and outputs with $\OMode{\texttt{red}}$. As a rule of
thumb, inputs are things which are supplied when checking the correctness of a
judgment, and outputs are things which are synthesized in the process. \fi

@@

  Intuitively, the general judgment $\GenJ{\Omega}{\mathcal{J}}$ is evident
  when the family of judgments $\mathcal{J}$ shall forevermore have a section
  (i.e.\ for any substitution $\Member{\omega}{\square\Omega}$,
  $\mathcal{J}(\omega)$ shall be evident). Recall that metavariables have
  \emph{valences} rather than sorts; as such, the general judgment allows us to
  quantify over terms of higher type.  However, for clarity we will often write
  $\GenJ{\MV{m}:\sigma}{\mathcal{J}}$ rather than
  $\GenJ{\MV{m}:\MkValence{\cdot}{\cdot}{\sigma}}{\mathcal{J}}$; likewise, in
  the body of $\mathcal{J}$ we will simply write $\MV{m}$ rather than the more
  proper $\MApp{\MV{m}}{\cdot}{\cdot}$.

@@

  In other words, the hypothetical judgment
  $\HypJ{\mathcal{J}_2}{\mathcal{J}_1}$ is evident when there shall forevermore
  be an effective transformation of evidence for the antecedent into evidence
  for the consequent. It is important to keep in mind that this is a
  \emph{semantic} consequence and therefore expresses admissibilities, in
  contrast to the \emph{logical} consequence which expresses
  derivabilities~\cite{harper:2016}. We will not have any need for the latter.
\end{definition}

@@

We will require two judgments, defined mutually: $\RefinesCtx{\Phi}{\Upsilon}$,
which means that $\Phi$ is a symbol context which refines the symbol context
$\Upsilon$ (presupposing $\IsSymCtx{\Upsilon}$), and parametric refinement
$\Refines{\Phi}{\phi}{\tau}$ (which shall presuppose
$\RefinesCtx{\Phi}{\Upsilon}$ and $\IsSort{\tau}$).

@@

Two refinements are equal when they denote the same PER.\@That is, will define
the refinement equality judgment $\EqRefines{\Phi}{\phi}{\psi}{\tau}$ as
follows, presupposing $\RefinesCtx{\Phi}{\Upsilon}$,
$\Refines{\Phi}{\phi}{\tau}$ and $\Refines{\Phi}{\psi}{\tau}$:
\begin{definition}
  To know $\EqRefines{\Phi}{\phi}{\psi}{\tau}$ is to know, for any substitution
  $\Of{\rho}{\Upsilon\hookrightarrow\OMode{\Upsilon'}}$, and for any $M, N$
  such that $\IsAbt{\cdot}{\Upsilon'}{\cdot}{M}{\tau}$ and
  $\IsAbt{\cdot}{\Upsilon'}{\cdot}{N}{\tau}$,
  $\IMode{M}\sim^\rho_\phi\IMode{N}$ if and only if
  $\IMode{M}\sim^\rho_\psi\IMode{N}$.
\end{definition}
